A Mechanised Library for Generic (Co)Programming

### 2024-02-28 paper/paper.tex
working on Section 3.

In order to abstract recursion patters we need to be able to talk about functors
which have the suitable fixed-point properties, i.e. those who have an initial
algebra. Instances of these functors are the polynomial ones.  A common approach
to construct such functors is to use \emph{container}~\cite{AbbottAG05}.  In
this section we introduce the basics of polynomial functors as containers
(Section~\ref{sec:containers}).  However, reasoning about containers equality
will require us to consider both functional extensionality and heterogeneous
equality. As stated in previous sections~\ref{sec:intro}, this is rather
problematic since a stric requirement for this work is that we should remain
axiom-free. For this purpose in Section~\ref{sec:types-morphisms} we introduce
a custom equivalence relation on types in order to avoid using the functional
extensionality axiom.

\subsection{Functors and Containers}
\label{sec:containers}
A container is defined by a type of \emph{shapes} and a \emph{family of position
types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
An \emph{extension} of this container is a functor defined as follows:
\begin{coqcode}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms. Generally, any functor that is isomorphic
to a container extension is called a \emph{container functor}. However, in our
work we will call \emph{container functor} the functors defined by a specific
container extension.

@@

Finally we can define the type of containers we need for performing Coq
extraction correctly. First, containers are defined as pairs of shape and
position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean and lift booleans to propositions similarly to the implementation provided by
SSReflect~\cite{GonthierL09}.
For simplicity, we define the dependent pair of valid positions as a record,
which will be extracted to a singleton type with the validity proof erased by
Coq's code extraction mechanism.
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{coqcode}
Then, we define \emph{container extensions} as follows:
\begin{coqcode}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{coqcode}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{coqcode}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}

>>>>

In order to abstract recursion patters we need to be able to abstract away from
the particular shape of the data.  This is achieved by introducing the concept
of \emph{functors} which have the suitable fixed-point properties, i.e. those
who have a initial algebras. A common approach to construct such functors is to
use \emph{container}~\cite{AbbottAG05} which is what we introduce in this
Section~\ref{sec:containers}.  However, reasoning about containers equality will
require us to consider both functional extensionality and heterogeneous
equality. To avoid this in Section~\ref{sec:types-morphisms} we introduce a
custom equivalence relation on types.

\subsection{Functors and Containers}
\label{sec:containers}
To abstract away the shape of a particular data structure we use functors.
These are functions \coq{F : Type -> Type} which additionally have a
map \coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
functor represents a container for abstract data that can be manipulated without
changing the outer structure. For example, the type \coq{List : Type -> Type} is
a functor and \coq{List A} is the type of lists over elements of type \coq{A}
with the obvious \coq{fmap} function recursive over a list and applying the
function \coq{A -> B} for each element of the list.
Additionally the type of lists arises as the least fixed-point for the
functor \coq{F X = unit + (A * X) }.

In general not every functor has a fixed-point, so we are going to restrict
ourselves to polynomial functors as represented by containers.  A container is
defined by a type of \emph{shapes} and a \emph{family of position types} indexed
by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
In our previuos example, we would define \coq{Shape} as \coq{unit + A}
indicating there are two constructors in the data type and one of these contains
a piece of data of type \coq{A}. Moreover, we would define the positions
as \coq{Pos (inl tt) = Empty_set} and \coq{Pos (inr a) = unit} indicating that
the first constructor does not have any recursive type variables and the second
constructor has one type variable.

Therefore an \emph{extension} of this container is a functor defined as follows:
\begin{coqcode}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms. Note that for \coq{Pos (inl tt)} we have a
type of functions \coq{Empty_set -> X} which is only inhabited by the trivial
map from the empty type and for \coq{Pos (inr a)} we have the type of
functions \coq{unit -> X} which has cardinality \coq{unit}.

Generally, any functor that is isomorphic
to a container extension is called a \emph{container functor}. However, in our
work we will call \emph{container functor} the functors defined by a specific
container extension.

@@

Finally we can define the type of containers as follows:
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{coqcode}
where \coq{valid} is the First, containers are defined as pairs of shape and
position types that have a \emph{decidable} validity predicate. We represent
decidable predicates as functions that return a boolean and lift booleans to
propositions similarly to the implementation provided by
SSReflect~\cite{GonthierL09}.  For simplicity, we define the dependent pair of
valid positions as a record, which will be extracted to a singleton type with
the validity proof erased by Coq's code extraction mechanism.

Then, we define \emph{container extensions} as follows:
\begin{coqcode}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{coqcode}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{coqcode}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}
